Adaptive Safe Strong Screening for Efficient Lasso-Type Problems Optimization over the Solution Path

### 2021-01-08 main.tex
Updates from Overleaf

Safe screening rules can be implemented in one of two ways. Sequential rules are recalculated at each $\lambda$ value along the path using the previous solution, so they require heavy computation but are relatively powerful. One-step rules are only calculated once, at the beginning of the path. As a result, they require far less computational burden but they lose efficacy as the thus are faster but they cannot remain powerful along the path for long because they don't utilize previous solution.

In this paper, we propose an adaptive safe screening scheme. We begin by calculating a safe rule and applying it to multiple solution steps along the path.  However, unlike the one-step implementation, we monitor the power of this rule and once it is no longer effective, construct a new safe rule.  This process is repeated along the path, waiting until it is beneficial before investing in a new safe rule calculation. This scheme avoids disadvantages of both sequential rules and one-step rules. Reusing the same solution when the safe rule is still effective avoids unnecessary computations, while periodically updating the rules ensures that they maintain power across the entire path. We show that this adaptive approach can greatly reduce the computation time and outperforms other methods in all scenarios. Following the hybrid approaches, we also combined our approach with strong rules to form adaptive hybrid rules. Our method is general, and capable of utilizing any sequential safe rule, and thus readily extended to other lasso type models for which safe rules exist. Here, we illustrate its extensions to the logistic regression loss function and to the group lasso penalty.

@@

In convex optimization, the Karush-Kuhn-Tucker (KKT) conditions are both necessary and sufficient for any optimal solution.  Screening rules take advantage of these KKT conditions in order to discard inactive coefficients.  For the lasso, the KKT conditions governing the optimal solution $\hat{\beta}(\lambda)$ are given by
\begin{equation}
  \label{eq:kkt}
  \begin{aligned}
    x_j^Tr(\lambda)/n &= \lambda \, \sign \, \hat{\beta}(\lambda)_j & & \textrm{if } \hat{\beta}(\lambda)_j\neq 0,\\
    |x_j^Tr(\lambda)/n| &\leq \lambda & & \textrm{if } \hat{\beta}(\lambda)_j= 0.
  \end{aligned}
\end{equation}
If the residuals were known, we could safely conclude that $\hat{\beta}(\lambda)_j=0$ by verifying that the second condition holds as a strict inequality.  The complication, of course, is that the residuals are not known until after we have solved for all the coefficients.  However, when fitting a solution path, residuals at previous $\lambda$ values provide information about residuals at the next $\lambda$ value and thus may be used for screening. Note that if we have $\lambda_0=\max_j|x_j^Ty/n|$ and $\hat{\beta}(\lambda_0)=0$, then $r(\lambda_0)=y$ satisfies the KKT conditions and thus $\hat{\beta}(\lambda_0)=0$ is the optimal solution.

@@

After computing the solution, a post-convergence KKT conditions check is performed. Because the KKT conditions are sufficient for the solution to be optimal, if there is no violation, the solution to the smaller model will be exactly the same as if we solve the model with all features. If there are violations, corresponding features will be added to the model and this procedure is repeated. Because features enter the model relatively infrequently, typically only a small number of repetitions are needed, thereby providing a substantial speed up compared to solving the full lasso problem every time.

@@

The biggest problem of SSR is that the ``unit-slope'' assumption may not hold and thus discarded features are not guaranteed to have zero coefficients. As with AC, a post-convergence KKT check is therefore needed for the discarded features to make sure their coefficients are truly zero. If violations occur, we again need to refit the model, adding the violating features and repeating the post-convergence check. However, previous studies have shown that violations are rare, so the additional $O(np)$ cost for repeating a KKT check hardly ever arises.

@@

\begin{theorem}[BEDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$ and $x_*=argmax_{x_j}|x_j^Ty/n|$. Under \eqref{eq:std} and for any $\lambda\in(0,\lambda_0]$, we have $\hat{\beta}_j(\lambda)=0$ if
    \begin{equation}
        |(\lambda_0+\lambda)x_j^Ty-(\lambda_0-\lambda)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda-(\lambda_0-\lambda)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
\end{theorem}

\begin{theorem}[SEDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition \eqref{eq:std}
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then $\hat{\beta}_j(\lambda_{k+1})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+1}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+1})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+1}-(\lambda_k-\lambda_{k+1})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, SEDPP rule reduces to BEDPP rule. $\hat{\beta}_j(\lambda_1)=0$ if (9) holds with $\lambda=\lambda_1$.
    \end{enumerate}
\end{theorem}

In BEDPP rule (4), the most costly parts of computation that take $O(np)$ time are the computation of $\{x_j^Ty\}_{j=1}^p$ and $\{x_j^Tx_*\}_{j=1}^p$ ($x_*^Ty$ is an element of $\{x_j^Ty\}_{j=1}^p$). These quantities do not depend on $\lambda_k$ and thus only need to be computed once at the beginning and stored. After this step, other computation of the rule takes almost zero time, so BEDPP rule can be viewed as an one-step rule with $O(np)$ cost.

In contrast, the SEDPP rule (5) requires $\{x_j^Tr(\lambda_k)\}_{j=1}^p$, which must be recomputed at every $\lam$.  This is the primary source of computational burden, as the other terms in the rule are $\{x_j^Ty\}_{j=1}^p$, which only needs to be calculated once, and $x_j^T\hat{y}(\lambda_k)$, which can be computed as $x_j^Ty-x_j^Tr(\lambda_k)$. Thus, the sequential safe rule SEDPP requires $O(npK)$ operations.

@@

From the SEDPP rule (6) in Theorem 2.2, we can do screening at $\lambda_{k+1}$ given the solution at $\lambda_k$. Now consider we are solving the lasso problem along the path $\lambda_0,...,\lambda_k,\lambda_{k+2},...,\lambda_K$ instead. Using the SEDPP rule, we can do screening at $\lambda_{k+2}$ as well, given the solution at $\lambda_k$, which is considered the previous solution now. Similarly, with solution at $\lambda_k$ as the reference, we can keep screening for a number of $\lambda$-values: $\lambda_{k+1},\lambda_{k+2},...$ until it is beneficial to update the solution at some $\lambda_{K+B}$ as the new reference and in this way all the discarded features are guaranteed to be zero. Adaptive EDPP is still a safe rule, but it is neither sequential nor one-step.

@@

\begin{theorem}[Adaptive EDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$ and the last updated reference point is $\lambda_k$. Under condition \eqref{eq:std}
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some next update point $k+B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some next update point $k+B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in one update $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (7) or (8) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$, $x_j^Tx_*$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ and $x_j^Tx_*$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the update and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that update. These two computations take $O(np)$ time and the rest computation in (7) and (8) take at most a lower order of $O(p)$ time.

@@

This combination is not trivial. For ordinary SSR, the quantities $x_j^Tr(\lambda_k)$ used in (4) is already computed in the previous post-convergence check in (3). However, combined with safe rules, previous post-convergence check only needs to check features that are not discarded at the previous $\lambda$-value. Thus, for features that are discarded at the previous $\lambda$-value but not at the current $\lambda$-value, their corresponding $x_j^Tr(\lambda_k)$ have not been computed yet and thus cannot be used without additional cost. Similar to the idea of BASIL \citep{qian2019fast}, we can also do the strong rule screening adaptively by using the latest $x_j^Tr(\lambda_k)$ available, which is computed by adaptive EDPP rule at the latest update, to reduce the number of required computations.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The rest calculations, those with $O(p)$ complexity for example, are not only faster in order, but also consume less memory and thus usually do no need out-of-core computing. Calculation of ${x_j^Ty}_{j=1}^p$ is needed only once for the whole solution path, so it would be neglected too. We will assume that strong rule screening is always powerful in the sense that $|\mathcal{R}_b|=o(p)\approx0$, but safe rule screening may not. We also assume that post-convergence check will mostly pass in the first iteration and thus is only needed approximately once per $\lambda$-value.

@@

\begin{equation}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{k+b}|}{B}.
\end{equation}

@@

This example also shows the timing of updates should be decided carefully and adaptively. For the first 10 $\lambda$-values, the safe rule stays powerful, which means the old reference is still exploitable. We may want delay the update to reduce average safe screening cost without harming the KKT condition checking cost much. For the last 20 $\lambda$-values however, the screening power drops to 0 quickly, which means the reference becomes useless quickly. Updating more frequently can greatly reduce the KKT condition checking cost with relatively small increase in safe screening cost. Also, the performance of adaptive EDPP rule also depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. These suggest that simple updating rules like periodic updating cannot not be consistently efficient and we want a method that can automatically determine whether to update given current performance.

@@

From previous example, we assume the average cost is a convex function of distance from last update $B$, which decreases first and then after some optimal value increases again. If we want to minimize the average cost per $\lambda$-value, we need to update when using the old reference for one more $\lambda$-value will increase the average cost. That is, we should update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B+1)>\bar{T}_k(B)$. For practice, we will update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B)>\bar{T}_k(B-1)$ instead, because $\bar{T}_k(B+1)$ cannot be calculated without doing screening at $\lambda_{k+B+1}$. These two conditions should be close and thus this substitution should also be close to optimal. Combined with (9), the latter condition will approximately be:

\begin{equation}
    (B-1)|\mathcal{S}_{k+B}|-\sum_{b=1}^{B-1}|\mathcal{S}_{k+b}|>p.
\end{equation}

This updating condition is a greedy condition that only considers the cost up to the current $\lambda$ value. In practice, we cannot know about future $\lambda$ values so this is the most reasonable condition. In fact, it provides a robust strategy towards three most common scenarios. First, if the screening power stays high, meaning $\mathcal{S}_{k+b}$ is small for all $b$, then the left-hand-side of (10) will be small and thus at this point we will not update to further exploit the last update. Second, if the screening power has been high for a while but dropped suddenly, meaning $\mathcal{S}_{k+B}$ is large but $\mathcal{S}_{k+b}$ are mostly small for $b=1,2,...,B-1$, then the left-hand-side of (10) will be large and we will update the useless old reference to a new helpful one. Last, if the screening power drops dramatically right after the last update, meaning $\mathcal{S}_{k+b}$ are close to $p$ for $b=2,3,...,B$, then the left-hand-side of (10) will be approximately $p-|\mathcal{S}_1|<p$. We will not update although old reference is useless, because if we did, the new update would become useless quickly as well.

@@

\begin{equation}
    \hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}-\frac{1}{n}\sum_{i=1}^n\{y_i\log\pi_i(\beta)+(1-y_i)\log(1-\pi_i(\beta))\}+\lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta+\beta_0\mathbf{1}$ is the linear function. $X$ are still the $n\times p$ standardized feature matrix but in this model the $y\in\{0,1\}^n$ is an unstandardized binary response vector. $\beta$ is still the $p\times1$ coefficient vector but we also need a $\beta_0$ as the intercept.

We can use SSR to do the strong rule screening with a small change according to the KKT condition. The residual vector now is $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$. Then the SSR discards the $j$-th feature at $\lambda_{k+1}$ if (4) holds.

@@

\begin{theorem}[Adaptive Slores]
For the sparse logistic regression model (15), let $\lambda_0:=\max_j|x_j^Ty/n|$, $\Tilde{y}=2y-1\in\{-1,1\}^n$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. $\forall\theta\in(0,1)^n$, let

@@

\begin{equation}
    \hat{\beta}(\lambda) = \underset{\beta\in \mathbb{R}^p}{\mathrm{argmin}}\frac{1}{2n}\bigg|\bigg|y-\sum_{g=1}^GX_g\beta_g\bigg|\bigg|_2^2+\lambda\sum_{g=1}^G\sqrt{p_g}||\beta_g||_2,
\end{equation}
where $\beta=(\beta_1^T,...,\beta_G^T)^T$, $X_g$ is a $n\times p_g$ sub-matrix consisting of columns of $X$ corresponding to features in group $g$, $g=1,2,...,G$. Besides standardization in \eqref{eq:std}, we apply an addition level of standardization at the group level\citep{breheny2015group}:

\begin{equation}
    X_g^TX_g=nI_{p_g\times p_g},\qquad g=1,2,...,G.
\end{equation}

@@

There are several atypical cases when AHR's advantage over other methods is not as significant: \eqref{eq:lasso} $p$ or $n$ is small. This corresponds a small data set and computation will be less challenging. (2) $\beta_{max}$ is close to zero. This means all the nonzero coefficients are close to zero and are hard to be distinguished from zero coefficients. This can cause problem for any statistical learning methods. (3) $s$ is large. In this case, the underlying model that generate the data set is not sparse in the sense that proportion of nonzero coefficients is not small. Lasso model would not be the optimal model to analyse such an data set. (4) $L$ is small. In this case, only a coarse grid of $\lambda$-values are considered. The $\lambda$-value selected in this case may be far from optimal.

>>>>

Safe screening rules can be implemented in one of two ways. Sequential rules do screening at each $\lambda$ value along the path using the previous solution as the reference, so they require heavy computation but are relatively powerful. One-step only use the solution at the beginning of the path as the reference. As a result, they require far less computational burden but they lose efficacy as the thus are faster but they cannot remain powerful along the path for long because they don't utilize previous solution.

In this paper, we propose an adaptive safe screening scheme. We begin choosing a solution as the reference for the safe rule and applying it to multiple solution steps along the path. However, unlike the one-step implementation, we monitor the power of the rule with this reference and once it is no longer effective, update the reference of the safe rule to the latest solution.  This process is repeated along the path, waiting until it is beneficial before investing in a new safe rule reference. This scheme avoids disadvantages of both sequential rules and one-step rules. Reusing the same solution as reference when the safe rule is still effective avoids unnecessary computations, while periodically updating the reference ensures that they maintain power across the entire path. We show that this adaptive approach can greatly reduce the computation time and outperforms other methods in all scenarios. Following the hybrid approaches, we also combined our approach with strong rules to form adaptive hybrid rules. Our method is general, and capable of utilizing any sequential safe rule, and thus readily extended to other lasso type models for which safe rules exist. Here, we illustrate its extensions to the logistic regression loss function and to the group lasso penalty.

@@

In convex optimization, the Karush-Kuhn-Tucker (KKT) conditions are both necessary and sufficient for any optimal solution.  Screening rules take advantage of these KKT conditions in order to discard inactive coefficients.  For the lasso, the KKT conditions governing the optimal solution $\hat{\beta}(\lambda)$ are given by
\begin{equation}
  \label{eq:kkt}
  \begin{aligned}
    x_j^Tr(\lambda)/n &= \lambda \, \sign \, \hat{\beta}_j(\lambda) & & \textrm{if } \hat{\beta}_j(\lambda)\neq 0,\\
    |x_j^Tr(\lambda)/n| &\leq \lambda & & \textrm{if } \hat{\beta}_j(\lambda)= 0.
  \end{aligned}
\end{equation}
If the residuals were known, we could safely conclude that $\hat{\beta}_j(\lambda)=0$ by verifying that the second condition holds as a strict inequality.  The complication, of course, is that the residuals are not known until after we have solved for all the coefficients.  However, when fitting a solution path, residuals at previous $\lambda$ values provide information about residuals at the next $\lambda$ value and thus may be used for screening. Note that if we have $\lambda_0=\max_j|x_j^Ty/n|$ and $\hat{\beta}(\lambda_0)=0$, then $r(\lambda_0)=y$ satisfies the KKT conditions and thus $\hat{\beta}(\lambda_0)=0$ is the optimal solution.

@@

After computing the solution, a post-convergence KKT conditions check is performed. Because the KKT conditions are sufficient for the solution to be optimal, if there is no violation, the solution to the smaller model will be exactly the same as if we solve the model with all features. If there are violations, corresponding features will be added to the model and this procedure is repeated. There are a finite number of features that can be added, so the procedure will end after finite repetitions. Because features enter the model relatively infrequently, typically only a small number of repetitions are needed, thereby providing a substantial speed up compared to solving the full lasso problem every time.

@@

The biggest problem of SSR is that the ``unit-slope'' assumption may not hold and thus discarded features are not guaranteed to have zero coefficients. As with AC, a post-convergence KKT check is therefore needed for the discarded features to make sure their coefficients are truly zero. If violations occur, we again need to refit the model, adding the violating features and repeating the post-convergence check. Similar to the procedure in AC, there are a finite number of features that can be added, so the procedure will end after finite repetitions. Moreover, previous studies have shown that violations are rare, so the additional $O(np)$ cost for repeating a KKT check hardly ever arises.

@@

\begin{theorem}[BEDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$ and $x_*=argmax_{x_j}|x_j^Ty/n|$. Under \eqref{eq:std} and for any $\lambda\in(0,\lambda_0]$, we have $\hat{\beta}_j(\lambda)=0$ if
    \begin{equation}
        \label{eq:bedpp}
        |(\lambda_0+\lambda)x_j^Ty-(\lambda_0-\lambda)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda-(\lambda_0-\lambda)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
\end{theorem}

\begin{theorem}[SEDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition \eqref{eq:std}
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then $\hat{\beta}_j(\lambda_{k+1})=0$ if
        \begin{equation}
            \label{eq:sedpp}
            \begin{split}
                &\left|2\lambda_{k+1}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+1})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+1}-(\lambda_k-\lambda_{k+1})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, SEDPP rule reduces to BEDPP rule. $\hat{\beta}_j(\lambda_1)=0$ if \eqref{eq:bedpp} holds with $\lambda=\lambda_1$.
    \end{enumerate}
\end{theorem}

In BEDPP rule \eqref{eq:bedpp}, the most costly parts of computation that take $O(np)$ time are the computation of $\{x_j^Ty\}_{j=1}^p$ and $\{x_j^Tx_*\}_{j=1}^p$ ($x_*^Ty$ is an element of $\{x_j^Ty\}_{j=1}^p$). These quantities do not depend on $\lambda_k$ and thus only need to be computed once at the beginning and stored. After this step, other computation of the rule takes almost zero time, so BEDPP rule can be viewed as an one-step rule with $O(np)$ cost.

In contrast, the SEDPP rule \eqref{eq:sedpp} requires $\{x_j^Tr(\lambda_k)\}_{j=1}^p$, which must be recomputed at every $\lam$.  This is the primary source of computational burden, as the other terms in the rule are $\{x_j^Ty\}_{j=1}^p$, which only needs to be calculated once, and $x_j^T\hat{y}(\lambda_k)$, which can be computed as $x_j^Ty-x_j^Tr(\lambda_k)$. Thus, the sequential safe rule SEDPP requires $O(npK)$ operations.

@@

From the SEDPP rule \eqref{eq:sedpp} in Theorem 2.2, we can do screening at $\lambda_{k+1}$ using the solution at $\lambda_k$ as a reference. Now consider we are solving the lasso problem along the path $\lambda_0,...,\lambda_k,\lambda_{k+2},...,\lambda_K$ instead. Using the SEDPP rule, we can do screening at $\lambda_{k+2}$ using the solution at $\lambda_k$ as a reference as well, which is considered the previous solution now. Similarly, with solution at $\lambda_k$ as the reference, we can keep screening for a number of $\lambda$-values: $\lambda_{k+1},\lambda_{k+2},...$ until it is beneficial to update the reference to the newer solution at some $\lambda_{K+B}$ and in this way all the discarded features are guaranteed to be zero. Adaptive EDPP is still a safe rule, but it is neither sequential nor one-step.

@@

\begin{theorem}[Adaptive EDPP]
    For the lasso problem \eqref{eq:lasso}, let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$ and the last updated reference point is $\lambda_k$. Under condition \eqref{eq:std}
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some next update point $k+B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \label{eq:aedpp1}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some next update point $k+B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \label{eq:aedpp2}
            |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
        \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in one update $\lambda_{k+1},...,\lambda_{k+B}$, the computation in \eqref{eq:aedpp1} or \eqref{eq:aedpp2} is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$, $x_j^Tx_*$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ and $x_j^Tx_*$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the update and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that update. These two computations take $O(np)$ time and the rest computation in \eqref{eq:aedpp1} or \eqref{eq:aedpp2} take at most a lower order of $O(p)$ time.

@@

This combination is not trivial. For ordinary SSR, the quantities $x_j^Tr(\lambda_k)$ used in \eqref{eq:ssr} is already computed in the previous post-convergence check in \eqref{eq:kkt}. However, combined with safe rules, previous post-convergence check only needs to check features that are not discarded at the previous $\lambda$-value. Thus, for features that are discarded at the previous $\lambda$-value but not at the current $\lambda$-value, their corresponding $x_j^Tr(\lambda_k)$ have not been computed yet and thus cannot be used without additional cost. Similar to the idea of BASIL \citep{qian2019fast}, we can also do the strong rule screening adaptively by using the latest $x_j^Tr(\lambda_k)$ available, which is computed by adaptive EDPP rule at the latest update, to reduce the number of required computations.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The rest calculations, those with $O(p)$ complexity for example, are not only faster in order, but also consume less memory and thus usually do no need out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path, so it would be neglected too. We will assume that strong rule screening is always powerful in the sense that $|\mathcal{R}_b|=o(p)\approx0$, but safe rule screening may not. We also assume that post-convergence check will mostly pass in the first iteration and thus is only needed approximately once per $\lambda$-value.

@@

\begin{equation}
    \label{eq:cost}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{k+b}|}{B}.
\end{equation}

@@

This example also shows the timing of updates should be decided carefully and adaptively. For the first 10 $\lambda$-values, the safe rule stays powerful, which means the old reference is still exploitable. We may want delay the update to reduce average safe screening cost without harming the KKT condition checking cost much. For the last 20 $\lambda$-values however, the screening power drops to 0 quickly, which means the reference becomes useless quickly. Updating more frequently can greatly reduce the KKT condition checking cost with relatively small increase in safe screening cost. Moreover, the performance of adaptive EDPP rule also depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. These suggest that simple updating rules like periodic updating cannot not be consistently efficient and we want a method that can automatically determine whether to update given current performance.

@@

From previous example, we assume the average cost is a convex function of distance from last update $B$, which decreases first and then after some optimal value increases again. If we want to minimize the average cost per $\lambda$-value, we need to update when using the old reference for one more $\lambda$-value will increase the average cost. That is, we should update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B+1)>\bar{T}_k(B)$. For practice, we will update the reference to $\lambda_{k+B}$ when $\bar{T}_k(B)>\bar{T}_k(B-1)$ instead, because $\bar{T}_k(B+1)$ cannot be calculated without doing screening at $\lambda_{k+B+1}$. These two conditions should be close and thus this substitution should also be close to optimal. Combined with \eqref{eq:cost}, the latter condition will approximately be:

\begin{equation}
    \label{eq:crit}
    (B-1)|\mathcal{S}_{k+B}|-\sum_{b=1}^{B-1}|\mathcal{S}_{k+b}|>p.
\end{equation}

This updating condition is a greedy condition that only considers the cost up to the current $\lambda$ value. In practice, we cannot know about future $\lambda$ values so this is the most reasonable condition. In fact, it provides a robust strategy towards three most common scenarios. First, if the screening power stays high, meaning $\mathcal{S}_{k+b}$ is small for all $b$, then the left-hand-side of \eqref{eq:crit} will be small and thus at this point we will not update to further exploit the last update. Second, if the screening power has been high for a while but dropped suddenly, meaning $\mathcal{S}_{k+B}$ is large but $\mathcal{S}_{k+b}$ are mostly small for $b=1,2,...,B-1$, then the left-hand-side of \eqref{eq:crit} will be large and we will update the useless old reference to a new helpful one. Last, if the screening power drops dramatically right after the last update, meaning $\mathcal{S}_{k+b}$ are close to $p$ for $b=2,3,...,B$, then the left-hand-side of \eqref{eq:crit} will be approximately $p-|\mathcal{S}_1|<p$. We will not update although old reference is useless, because if we did, the new update would become useless quickly as well.

@@

\begin{equation}
    \label{eq:logis}
    \hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}-\frac{1}{n}\sum_{i=1}^n\{y_i\log\pi_i(\beta)+(1-y_i)\log(1-\pi_i(\beta))\}+\lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta+\beta_0\mathbf{1}$ is the linear function. $X$ are still the $n\times p$ standardized feature matrix but in this model the $y\in\{0,1\}^n$ is an unstandardized binary response vector. $\beta$ is still the $p\times1$ coefficient vector but we also need a $\beta_0$ as the intercept.

We can use SSR to do the strong rule screening with a small change according to the KKT condition. The residual vector now is $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$. Then the SSR discards the $j$-th feature at $\lambda_{k+1}$ if \eqref{eq:ssr} holds.

@@

\begin{theorem}[Adaptive Slores]
For the sparse logistic regression model \eqref{eq:logis}, let $\lambda_0:=\max_j|x_j^Ty/n|$, $\Tilde{y}=2y-1\in\{-1,1\}^n$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. $\forall\theta\in(0,1)^n$, let

@@

\begin{equation}
    \label{eq:glasso}
    \hat{\beta}(\lambda) = \underset{\beta\in \mathbb{R}^p}{\mathrm{argmin}}\frac{1}{2n}\bigg|\bigg|y-\sum_{g=1}^GX_g\beta_g\bigg|\bigg|_2^2+\lambda\sum_{g=1}^G\sqrt{p_g}||\beta_g||_2,
\end{equation}
where $\beta=(\beta_1^T,...,\beta_G^T)^T$, $X_g$ is a $n\times p_g$ sub-matrix consisting of columns of $X$ corresponding to features in group $g$, $g=1,2,...,G$. Besides standardization in \eqref{eq:std}, we apply an addition level of standardization at the group level\citep{breheny2015group}:

\begin{equation}
    \label{eq:stdg}
    X_g^TX_g=nI_{p_g\times p_g},\qquad g=1,2,...,G.
\end{equation}

@@

There are several atypical cases when AHR's advantage over other methods is not as significant: (1) $p$ or $n$ is small. This corresponds a small data set and computation will be less challenging. (2) $\beta_{max}$ is close to zero. This means all the nonzero coefficients are close to zero and are hard to be distinguished from zero coefficients. This can cause problem for any statistical learning methods. (3) $s$ is large. In this case, the underlying model that generate the data set is not sparse in the sense that proportion of nonzero coefficients is not small. Lasso model would not be the optimal model to analyse such an data set. (4) $L$ is small. In this case, only a coarse grid of $\lambda$-values are considered. The $\lambda$-value selected in this case may be far from optimal.
